---
title: "605-HW15-Calculus 3"
author: "Michael Y."
date: "December 8, 2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    code_folding: show
    toc: yes
    toc_float: yes
    toc_depth: 3
    md_extensions: +grid_tables
  pdf_document:
    md_extensions: +grid_tables
    toc: yes
    toc_depth: 3
classoption: portrait
editor_options:
  chunk_output_type: inline
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
---

<style>
  .main-container {
    max-width: 1200px !important;
  }
</style>

---


```{r setup, eval=T, echo=F}
# Setup
knitr::opts_chunk$set(echo = TRUE,
                      fig.pos='H')
directory = "C:/Users/Michael/Dropbox/priv/CUNY/MSDS/201909-Fall/DATA605_Larry/20191208_Week15/"
knitr::opts_knit$set(root.dir = directory)

### Make the output wide enough
options(scipen = 999, digits=9, width=150)
```

```{r load, eval=T, message=F,warning=F, echo=F}
### load some libraries
library(pracma)
library(kableExtra)
```

\newpage
# HW15 - Calculus 3

## 1. Find the equation of the regression line for the given points.   
#### Round any final values to the nearest hundredth, if necessary.

```{r ex1, eval=TRUE}
# put the pairs into a list
points = list(
  c( 5.6,  8.8 ), 
  c( 6.3, 12.4 ), 
  c( 7,   14.8 ), 
  c( 7.7, 18.2 ), 
  c( 8.4, 20.8 )
  )
# Make the list into an array
points=simplify2array(points)
points

# get x
x=points[1,]
x

# get y
y=points[2,]
y

# regression
model1 =  lm(y~x)
model1
summary(model1)
beta=model1$coefficients
beta

intercept    =round(beta["(Intercept)"],2)
slope        =round(beta["x"],2)
eqn = paste("Regression line: y = ", intercept, "+", slope, "x")
eqn
plot(y~x, main=eqn)
abline(model1, col="red")
```

The formula for the regression line is $\boxed{`r paste("y = ", intercept, "+", slope, "x")`}$ .   


***
\newpage

## 2. Find all local maxima, local minima, and saddle points for the function given below. 

#### Write your answer(s) in the form ( x, y, z ). 
#### Separate multiple points with a comma.     


$f ( x, y ) = 24x - 6xy^2 - 8y^3$

To find the critical points, compute the first derivatives and set equal to zero:    

$f_x = \frac{\partial}{\partial x}(24x - 6xy^2 -8y^3) = 24-6y^2 = 0 \implies4 = y^2 \implies y \in \{-2,2\}$   .

$f_y = \frac{\partial}{\partial y}(24x - 6xy^2 -8y^3) = -12xy-24y^2 = 0$    .     

$xy =-2y^2 \implies (y=0) \vee (x=-2y)$ .    

From above, we know that $y \in \{-2,2\}$ so we can disregard the case that $y=0$.     

Therefore, $x \in \{4,-4\}$, i.e., $(x,y) \in \{(-4,2),(4,-2)\}$ .  

Evaluating the function at these points, we obtain:    

$f(-4,2) = 24\cdot(-4) - 6\cdot(-4)\cdot(2)^2 - 8\cdot(2)^3=-96+96-64=-64$     
$f(4,-2) = 24\cdot(4) - 6\cdot(4)\cdot(-2)^2 - 8\cdot(-2)^3=96-96+64=64$    

Therefore, the ***critical points*** are $\boxed{(x,y,z) \in \{(-4,2,-64),(4,-2,64)\}}$ .       

To determine whether these critical values are maxima, minima, or saddle points, we need to use the second partial derivative test:   

$f_{xx} = \frac{\partial(f_{x})}{\partial x} = \frac{\partial}{\partial x}(24-6y^2) = 0$ .     

$f_{yy} = \frac{\partial(f_{y})}{\partial y} = \frac{\partial}{\partial y}(-12xy-24y^2) = -12x-48y$  .

$f_{xy}  = \frac{\partial(f_{x})}{\partial y} = \frac{\partial}{\partial y} (24-6y^2) = -12y$
$f_{yx}  = \frac{\partial(f_{y})}{\partial x} = \frac{\partial}{\partial x} (-12xy-24y^2) = -12y = f_{xy}$ , as expected.

The discriminant $D(f(x,y)) = f_{xx}f_{yy} - f_{xy}f_{yx} = (0)(-12x-48y)-(-12y)^2=-144y^2$.    
At both critical points, $D(f(-4,2)) = D(f(4,-2)) = -144\cdot(\pm2)^2 = -576 < 0$ .

By the second partial derivative test, each of the two points $(x,y,z) \in \{(-4,2,-64),(4,-2,64)\}$ is a ***saddle point*** -- there are no maxima nor minima.




***
\newpage

## 3. A grocery store sells two brands of a product, the "house" brand and a "name" brand. 

The manager estimates that if she sells the "house" brand for $x$ dollars and the "name" brand for $y$ dollars, she will be able to sell   

* $81 - 21x + 17y$ units of the "house" brand and 
* $40 + 11x - 23y$ units of the "name" brand.

### Step 1. Find the revenue function $R ( x, y )$ .    

$$\begin{aligned}
R(x,y) 
&= (81- 21x + 17y)x + (40 + 11x - 23y)y \\
&=81x - 21x^2 + 17xy + 40y +11xy- 23y^2 \\
&= -21x^2  -23y^2 + 28xy + 81x + 40y \\
\end{aligned}$$


### Step 2. What is the revenue if she sells the "house" brand for $2.30 and the "name" brand for $4.10?    

##### Calculation functions:
```{r calc-revenue, eval=T}
calcHouseUnits <- function(x,y)
{
  HouseUnits = 81 - 21*x + 17*y
  return(c(HouseUnits))
}

calcNameUnits <- function(x,y)
{
  NameUnits = 40 + 11*x + -23*y
  return(c(NameUnits))
}

calcRevenue <- function(x,y)
{
    r = - 21*x^2  - 23*y^2 + 28*x*y + 81*x + 40*y
    return(c(r))
}

calcRevenueWithRounding <- function(x,y)
{
  intHouseUnits = round(calcHouseUnits(x,y),0)
  intNameUnits = round(calcNameUnits(x,y),0)

  HouseRevenue = round(x,2) * intHouseUnits
  NameRevenue = round(y,2) * intNameUnits
  TotalRevenue = HouseRevenue + NameRevenue
return(c(TotalRevenue))
}
```

\newpage
### Calculate revenue without requiring that the quantity of each item sold be an integer
```{r calcs1}
x=2.30
y=4.10

houseUnits=calcHouseUnits(x,y)
print(paste("Units of House brand sold: ", houseUnits, 
            "at price of ",x," each ",
            "--> revenue = ", x*houseUnits))

nameUnits = calcNameUnits(x,y)
print(paste("Units of name brand sold: ", nameUnits, 
            "at price of ",y," each ",
            "--> revenue = ", y*nameUnits))

print(paste("Total revenue: ", x*houseUnits + y*nameUnits))

checkRevenue = calcRevenue(x,y)
print(paste("Double-check : ", checkRevenue))
```

\newpage
### Re-calculate revenue, adding a requirement that the quantity of each item sold be an integer
```{r calcs2}
x=2.30
y=4.10

houseUnits=round(calcHouseUnits(x,y),0)
print(paste("Units of House brand sold: ", houseUnits, 
            "at price of ",x," each ",
            "--> revenue = ", x*houseUnits))

nameUnits = round(calcNameUnits(x,y),0)
print(paste("Units of name brand sold: ", nameUnits, 
            "at price of ",y," each ",
            "--> revenue = ", y*nameUnits))

print(paste("Total revenue: ", x*houseUnits + y*nameUnits))

checkRevenue = calcRevenueWithRounding(x,y)
print(paste("Double-check : ", checkRevenue))
```

***There is a problem with these figures*** -- because it is expecting that we would sell 

* `r houseUnits` units of the "house" brand, but 
* `r nameUnits` units of the "name" brand.  

The latter figure is ***negative.***   **This is not possible.**

So, let's restrict the quantities to be non-negative:

\newpage
### Functions with non-negative quantity restrictions
```{r calc-nonneg-revenue, eval=T}
calcNonNegHouseUnits <- function(x,y)
{
  HouseUnits = 81 - 21*x + 17*y
  return(max(0,HouseUnits))
}

calcNonNegNameUnits <- function(x,y)
{
  NameUnits = 40 + 11*x + -23*y
  return(max(0,NameUnits))
}

#calcRevenue <- function(x,y)
#{
#    r = - 21*x^2  - 23*y^2 + 28*x*y + 81*x + 40*y
#    return(c(r))
#}

calcNonNegRevenueWithRounding <- function(x,y)
{
  intNonNegHouseUnits = round(calcNonNegHouseUnits(x,y),0)
  intNonNegNameUnits = round(calcNonNegNameUnits(x,y),0)

  NonNegHouseRevenue = round(x,2) * intNonNegHouseUnits
  NonNegNameRevenue = round(y,2) * intNonNegNameUnits
  NonNegTotalRevenue = NonNegHouseRevenue + NonNegNameRevenue

return(c(NonNegTotalRevenue))
}
```

\newpage
### Calculate results with non-negative restrictions:
```{r calcs3}
x=2.30
y=4.10

NonNegHouseUnits=round(calcNonNegHouseUnits(x,y),0)
print(paste("Units of House brand sold: ", NonNegHouseUnits, 
            "at price of ",x," each ",
            "--> revenue = ", x*NonNegHouseUnits))

NonNegNameUnits = round(calcNonNegNameUnits(x,y),0)
print(paste("Units of name brand sold: ", NonNegNameUnits, 
            "at price of ",y," each ",
            "--> revenue = ", y*NonNegNameUnits))

print(paste("Total revenue: ", x*NonNegHouseUnits + y*NonNegNameUnits))

checkNonNegRevenue = calcNonNegRevenueWithRounding(x,y)
print(paste("Double-check : ", checkNonNegRevenue))
```

***Still there is a problem here:***    
Allowing us to override an otherwise negative quantity-sold with zero, while still setting a high nominal price for such item, would allow us to sell an arbitrary (i.e., infinite) quantity of the other item simply by making the price of the non-salable item arbitrarily high.   

Under such model, the result would be an arbitrarily large amount of sales from from the house brand with zero sales of (expensive) name brand.

***This indicates that this problem is not well-formulated.***

\newpage
### Revenue maximization?

The question doesn't ask about maximization of revenue (which one would have expected it to ask...)   

So, let's compute it anyhow...    

* $R_x = -42x +28y +81 = 0$      
* $R_y =  28x -46y +40 = 0$

So, we have a system of 2 simultaneous equations in 2 unknowns:

$$\begin{cases} 
\begin{aligned}
-42x +28y +81 &=0 \\ 
28x  -46y +40 &= 0 
\end{aligned}
\end{cases}$$

Multiplying the second equation by $\frac{3}{2}$  gives us    
$$\begin{cases} 
\begin{aligned}
-42x +28y +81 &=0 \\ 
42x  -69y +60 &= 0 
\end{aligned}
\end{cases}$$

Adding together gives us   

$-41y+141=0$ , so $y = \frac{141}{41}\approx 3.43902439$   .

From the above equations, 
$x = \frac{28y+81}{42}=\frac{28 \cdot\frac{141}{41}+81}{42} =\frac{7269}{1722}\approx 4.22125436$   .    

So, the critical point (which we assume will be a maximum) is (housePrice,namePrice)=$(x^*,y^*)=\left(\frac{7269}{1722},\frac{141}{41}\right)\approx(4.22125436,3.43902439)$


Of course, we could use R to perform the above calculations for us:   
$$\begin{aligned}
R_x&:        -42x +28y  &= -81 \\
R_y&: \quad   28x -46y  &= -41
\end{aligned}$$



\newpage
### Solve system of 2 equations using `rref` from `pracma`
```{r solve-eqns, eval=T}
Rx = c(-42,28,-81)
Ry = c(28,-46,-40)
eqns <- rbind(Rx,Ry)
eqns

result <- rref(eqns)
result

soln <- result[,3]
soln
```

\newpage
### Perform calculations on $(x^{*},y^{*}) =(`r soln["Rx"]` , `r soln["Ry"]`)$

```{r calcs4}
xstar=soln["Rx"]
ystar=soln["Ry"]



starhouseUnits=calcHouseUnits(xstar,ystar)
print(paste("Units of House brand sold: ", starhouseUnits, 
            "at price of ",xstar," each ",
            "--> revenue = ", xstar*starhouseUnits))

starnameUnits = calcNameUnits(xstar,ystar)
print(paste("Units of name brand sold: ", starnameUnits, 
            "at price of ",ystar," each ",
            "--> revenue = ", ystar*starnameUnits))


print(paste("Total revenue: ", xstar*starhouseUnits + ystar*starnameUnits))

starrevenue = calcRevenue(xstar,ystar)
starrevenue

```

##### The above indicates that we generate revenue of \$$`r starrevenue`$.


Of course, we should assume that 

* we can't sell fractional units,  and  
* each unit price should be rounded to cents.     

This introduces an "integer programming" problem, which can be considerably more difficult than a linear programming problem, because the result obtained from rounding the quantities to the closest integer may be inferior to rounding to the more distant integer.    

Ignoring such issue, we obtain a slightly smaller result:

\newpage
### Perform calculations on $(x^{*},y^{*}) =(`r round(xstar,2)` , `r round(ystar,2)`)$

```{r calcs5}
roundxstar=round(xstar,2)
roundystar=round(ystar,2)

intstarhouseUnits=round(calcHouseUnits(roundxstar,roundystar),0)
print(paste("Units of House brand sold: ", intstarhouseUnits, 
            "at price of ",roundxstar," each ",
            "--> revenue = ", roundxstar*intstarhouseUnits))

intstarnameUnits = round(calcNameUnits(roundxstar,roundystar),0)
print(paste("Units of name brand sold: ", intstarnameUnits, 
            "at price of ",roundystar," each ",
            "--> revenue = ", roundystar*intstarnameUnits))


print(paste("Total revenue: ", roundxstar*intstarhouseUnits + roundystar*intstarnameUnits))

starrevenue = calcRevenueWithRounding(roundxstar,roundystar)
starrevenue
```

##### The above indicates that we generate revenue of \$$`r starrevenue`$.


\newpage
### Confirm maximum

To verify that this critical point is in fact a maximum, consider the second derivative test:   

$R_x = -42x +28y +81 = 0$
$R_y =  28x -46y +40 = 0$


$R_{xx} = \frac{\partial(R_{x})}{\partial x} = \frac{\partial}{\partial x}(-42x +28y +81) = -42$ .     

$R_{yy} = \frac{\partial(R_{y})}{\partial y} = \frac{\partial}{\partial y}(28x -46y +40) = -46$  .

$R_{xy}  = \frac{\partial(R_{x})}{\partial y} = \frac{\partial}{\partial y} ( -42x +28y +81) = 28$
$R_{yx}  = \frac{\partial(R_{y})}{\partial x} = \frac{\partial}{\partial x} (28x -46y +40) = 28 = f_{xy}$ , as expected.

The discriminant 
$$\begin{aligned}
D(f(x,y)) 
= R_{xx}R_{yy} - R_{xy}R_{yx} 
&= (-42)(-46)-(28)^2\\
&=1932-784\\
&=1148\\
&>0
\end{aligned}$$.

Because

* $D(R(x,y))>0 \quad\forall (x,y)$ , and 
* $R_{xx}(x,y)<0 \quad\forall(x,y)$, 

the critical point (`r roundxstar`,`r roundystar`) is a ***maximum*** .   

### However, there is still one problem here:   

Generally a "house" brand is ***less expensive*** than a "name" brand.   

The result of this optimization is to select 

* a ***higher price*** (`r roundxstar`) for the ***house brand*** and 
* a ***lower price*** (`r roundystar`) for the ***name brand***.    

***This does not make sense***, and further suggests that the model is not well-formulated.   



***
\newpage

## 4. A company has a plant in Los Angeles and a plant in Denver. 

The firm is committed to produce a total of 96 units of a product each week.    

The total weekly cost is given by $C(x, y) = \frac{1}{6}x^2 +  \frac{1}{6}y^2 + 7x + 25y + 700$  , where     

* $x$ is the number of units produced in Los Angeles and 
* $y$ is the number of units produced in Denver.     

How many units should be produced in each plant to **minimize** the total weekly cost?     

First, note that $x+y=96$, so $y=96-x$ .    

Substituting,    

$$\begin{aligned}
C(x, y) &= \frac{1}{6}x^2 +  \frac{1}{6}y^2 + 7x + 25y + 700 \\
C(x, 96-x)&= \frac{1}{6}x^2 +  \frac{1}{6}(96-x)^2 + 7x + 25(96-x) + 700 \\
&= \frac{1}{6}x^2 +  \frac{1}{6}(96^2-192x+x^2) + 7x + (2400-25x) + 700 \\
&= \frac{1}{6}x^2 +  1536 -32x + \frac{1}{6}x^2 + 7x + (2400-25x) + 700 \\
&= \frac{1}{3}x^2 -50x  +4636 \\
&= \frac{x^2-150x+5625}{3} +2721 \\
&= \frac{(x-75)^2}{3}+2761
\end{aligned}$$


Next, compute the derivative $\frac{dC}{dx}$ and set equal to zero:

$C' = \frac{dC}{dx} = \frac{2(x-75)}{3}=\frac{2}{3}x-50=0 \\ \implies x=75 \\ \implies y=96-x=96-75=21$.

To confirm that this is a ***minimum***, we look at the second derivative test:

$C'' = \frac{d^2C}{dx^2} = \frac{2}{3} > 0$.   

Because the second derivative is positive, this confirms that the critical point is a minimum.   

This means that:      

* 75 units should be produced in Los Angeles
* 21 units should be produced in Denver

Under the above allocation, the total weekly cost is minimized at $C(75,21) = \$2,761$ .

***
\newpage

## 5. Evaluate the double integral ${\iint\limits_R {e^{8x + 3y} dA}}$ on the given region .
R: $2 \le x \le 4$ and $2 \le y \le 4$

Write your answer in exact form without decimals.

$$\begin{aligned}
\int\limits_{2}^4 \left[\int\limits_{2}^4 {e^{8x + 3y} dx}\right] dy 
&=\int\limits_{2}^4 \left[e^{3y}\int\limits_{2}^4 {e^{8x} dx}\right] dy \\
&=\int\limits_{2}^{4} e^{3y}\left[ \frac{e^{8x}}{8} \right]_{x=2}^{x=4} dy \\
&=\int\limits_{2}^{4} e^{3y}\left[ \frac{e^{32}-e^{16}}{8} \right] dy \\
&=\left[ \frac{e^{32}-e^{16}}{8} \right]\int\limits_{2}^{4} e^{3y} dy \\
&=\left[ \frac{e^{32}-e^{16}}{8} \right]\left[ \frac{e^{3y}}{3} \right]_{y=2}^{y=4} \\
&=\left[ \frac{e^{32}-e^{16}}{8} \right]\left[ \frac{e^{12}-e^{6}}{3} \right] \\
&=\left[ \frac{e^{32}e^{12}-e^{32}e^{6}-e^{16}e^{12}+e^{16}e^{6}}{24} \right] \\
&=\left[ \frac{e^{44}-e^{38}-e^{28}+e^{22}}{24} \right] \\ \\
&=\boxed{\frac{e^{22}}{24}\left[ {e^{22}-e^{16}-e^{6}+1} \right]} \\
\end{aligned}$$
















