---
title: "605-HW02-MatrixFactorization"
author: "Michael Y." 
date: "September 8, 2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    code_folding: show
    toc: yes
    toc_float: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
classoption: landscape
editor_options:
  chunk_output_type: inline
---
<style>
  .main-container {
    max-width: 1200px !important;
  }
</style>---

#### Setup
```{r setup, eval=T}
knitr::opts_chunk$set(echo = TRUE)
directory = "C:/Users/Michael/Dropbox/priv/CUNY/MSDS/201909-Fall/DATA605_Larry/20190908_Week02/"
knitr::opts_knit$set(root.dir = directory)
knitr::opts_knit$set(digits=6)

### Make the output wide enough
options(scipen = 999, digits=6, width=150)

### Load some libraries
library(tidyr)
library(dplyr)
library(kableExtra)
library(pracma)

```

#### Function to write Matrix, courtesy of Vinayak Patel :
``` {r writemtx, eval=T}
writeMatrix <- function(x) {  
  begin <- "\\begin{bmatrix}"  
  end   <- "\\end{bmatrix}"  
  X     <-    apply(x, 1, function(x) {
      paste(
          paste(x, collapse = "&"),
          "\\\\"
      )
    }
  )  
  paste(c(begin, X, end), 
        collapse = "")
  }
```

#### Function to write numerical Matrix, controlling decimals, adapted :
``` {r wnmtx, eval=T}
wnumMatrix <- function(x) {  
  begin <- "\\begin{bmatrix}"  
  end   <- "\\end{bmatrix}"  
  X     <-    apply(x, 1, function(x) {
      paste(
          paste(format(x, digits = options()$digits), collapse = "&"),
          "\\\\"
      )
    }
  )  
  paste(c(begin, X, end), 
        collapse = "")
  }
```


# Part 1 - (non)Commutivity of Matrix Transposes


## (1) Show that ${A^{T}A \neq AA^{T}}$ in General.  (Proof and Demonstration)

```{r rankA, eval=T, echo=F}
A = c(
 "A_{1,1}", "A_{1,2}", "A_{1,j}", "A_{1,n}", 
 "A_{2,1}", "A_{2,2}", "A_{2,j}", "A_{2,n}", 
 "A_{i,1}", "A_{i,2}", "A_{i,j}", "A_{i,n}", 
 "A_{n,1}", "A_{n,2}", "A_{n,j}", "A_{n,n}"
)
A=matrix(A,nrow=4,byrow=T)
A
AT = t(A)
AT
```
$$ A = `r writeMatrix(A)` $$
$$ A^T = `r writeMatrix(AT)` $$
$${ AA^{T}=  `r writeMatrix(A)`  `r writeMatrix(AT)` } =  \left[ \begin{matrix} 
\sum\limits_{ j=1 }^{ n }{ { A }_{ 1,j }^{ 2 } }  & \sum\limits_{ j=1 }^{ n }{ { A }_{ 1,j }{ A }_{ 2,j } }  & ... & \sum\limits_{ j=1 }^{ n }{ { A }_{ 1,j }{ A }_{ n,j } }  \\ 
\sum\limits_{ j=1 }^{ n }{ { A }_{ 1,j }{ A }_{ 2,j } }  & \sum\limits_{ j=1 }^{ n }{ { A }_{ 2,j }^{ 2 } }  & ... & \sum\limits_{ j=1 }^{ n }{ { A }_{ 2,j }{ A }_{ n,j } }  \\ 
... & ... & ... & ... \\ 
\sum\limits_{ j=1 }^{ n }{ { A }_{ 1,j }{ A }_{ n,j } }  & \sum\limits_{ j=1 }^{ n }{ { A }_{ 2,j }{ A }_{ n,j } }  & ... & \sum\limits_{ j=1 }^{ n }{ { A }_{ n,j }^{ 2 } }  
\end{matrix} \right]  $$


$${ A^TA = `r writeMatrix(AT)` `r writeMatrix(A)` } = \left[ \begin{matrix} 
\sum\limits_{ i=1 }^{ n }{ { A }_{ i,1 }^{ 2 } }  & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,1 }{ A }_{ i,2 } }  & ... & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,1 }{ A }_{ i,n } }  \\ 
\sum\limits_{ i=1 }^{ n }{ { A }_{ i,1 }{ A }_{ i,2 } }  & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,2 }^{ 2 } }  & ... & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,2 }{ A }_{ i,n } }  \\ 
... & ... & ... & ... \\ 
\sum\limits_{ i=1 }^{ n }{ { A }_{ i,1 }{ A }_{ i,n } }  & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,2 }{ A }_{ i,n } }  & ... & \sum\limits_{ i=1 }^{ n }{ { A }_{ i,n }^{ 2 } }  
\end{matrix} \right] $$


For the above to be true, we would need to have the above summations to be elementwise equal between the top and bottom matrices.     
This is usually not the case.   

### COUNTEREXAMPLE:    

```{r counterexample, eval=T}
Q = c(1,2,3,6,4,5,7,8,9) 
Q=matrix(Q,3,3,T) 
Q
# Transpose
QT = t(Q)
QT
# Q * QT
QQT = Q %*% QT
QQT
# QT * Q
QTQ = QT %*% Q
QTQ
# Are the entries equal?
QQT == QTQ

```

As a simple counterexample, consider:      


$$ Q = `r writeMatrix(Q)` ; Q^{T} = `r writeMatrix(QT)`$$
$$ QQ^{T} = `r writeMatrix(Q)` `r writeMatrix(QT)` = `r writeMatrix(QQT)`$$
$$ Q^{T}Q = `r writeMatrix(QT)` `r writeMatrix(Q)` = `r writeMatrix(QTQ)`$$
Clearly, ${QQ^T \neq Q^TQ}$   .  



***  

## (2) For a special type of square matrix,  ${ A^{ T }A = AA^{ T } }$   
#### Under what conditions could this be true? (Hint: The Identity matrix I is an example of such a matrix).



This is true for **Permutation** matrices because 

$$(PP^T)_{ij} = \sum_{k=1}^n P_{ik} P^T_{kj} = \sum_{k=1}^n P_{ik} P_{jk}$$
and   

$$\sum_{k=1}^n P_{ik} P_{jk} = \begin{cases} 1 & \text{if } i = j \\ 0 & \text{otherwise} \end{cases}$$
so $$PP^T = I$$ .

### PERMUTATION EXAMPLE:  
```{r permutation, eval=T}
P = c(
0,1,0,0,0,
0,0,0,1,0,
1,0,0,0,0,
0,0,1,0,0,
0,0,0,0,1
)
P=matrix(P,5,5,T)
P
PT=t(P)
PT
PPT = P %*% PT
PPT
PTP = PT %*% P
PTP
PPT==PTP
```

For example, consider:    

$$ P = `r writeMatrix(P)` ; P^{T} = `r writeMatrix(PT)`$$
$$ PP^{T} = `r writeMatrix(P)` `r writeMatrix(PT)` = `r writeMatrix(PPT)` = I_5$$
$$ P^{T}P = `r writeMatrix(PT)` `r writeMatrix(P)` = `r writeMatrix(PTP)` =I_5$$

***


Obviously commutativity is true for **symmetric** matrices because, by definition,  ${A = A^{T}}$, so ${AA^T}=AA=A^TA$   .      

### SYMMETRIC EXAMPLE:
```{r symmetric, eval=T}
S = c(
  1,2,4,
  2,3,5,
  4,5,6
  )
S = matrix(S,3,3,T)
S
ST=t(S)
ST
SST = S %*% ST
SST
STS = ST %*% S
STS
SST == STS
```


For example, consider:    

$$ S = `r writeMatrix(S)` ; S^{T} = `r writeMatrix(ST)`$$
$$ SS^{T} = `r writeMatrix(S)` `r writeMatrix(ST)` = `r writeMatrix(SST)`$$
$$ S^{T}S = `r writeMatrix(ST)` `r writeMatrix(S)` = `r writeMatrix(STS)`$$



***


Also this is true for **orthogonal** matrices because, by definition,   ${A^{T} = A^{-1}det(A)}$ ,   
so $${ AA^{T} = AA^{-1}*det(A)=I*{det(A)}=A^{-1}A*det(A)=A^{T}A}$$    .    
     
    
### ORTHOGONAL EXAMPLE:      
```{r orthogonal,eval=T}

X = c(
  1,-1,
  1,1
  )
X=matrix(X,2,2,T)
X
# Compute transpose of X
XT=t(X)
XT
# Compute inverse of X
Xinv = inv(X)
Xinv
# Compute determinant of X
Xdet = det(X)
Xdet
# Compute XXT
XXT = X %*% XT
XXT
# Compute XTX
XTX = XT %*% X
XTX
# Do the entries match?
XXT == XTX
```

  
For example, consider **orthogonal** matrix ${X = `r writeMatrix(X)`}$ which has transpose ${X^{T} = `r writeMatrix(XT)`}$  . Here,        
${XX^{T}= `r writeMatrix(X)` `r writeMatrix(XT)` = `r writeMatrix(XXT)`}$     
${X^{T}X= `r writeMatrix(XT)` `r writeMatrix(X)` = `r writeMatrix(XTX)`}$

Note that we can make the above matrix **orthonormal** by dividing it by the square root of its determinant, which here is 2.    

    
### ORTHONORMAL EXAMPLE:        
```{r orthonormal,eval=T}

# make orthonormal
O = X / sqrt(det(X))
O
# compute transpose
OT=t(O)
OT
# Compute inverse
Oinv = inv(O)
Oinv
# does transpose "equal" inverse (up to machine precision?)
OT - Oinv
abs(OT-Oinv) < 1e-12
# Compute OOT
OOT = O %*% OT
OOT
# Compute OTO
OTO = OT %*% O
OTO
# do the entries match?
OOT == OTO
```


For example, consider **orthonormal** matrix ${O = `r wnumMatrix(O)`}$ which has transpose ${O^{T} = `r wnumMatrix(OT)`}$  .   
Here,       
${OO^{T}= `r wnumMatrix(O)` `r wnumMatrix(OT)` = `r wnumMatrix(OOT)`} = I_2$     
   
${O^{T}O= `r wnumMatrix(OT)` `r wnumMatrix(O)` = `r wnumMatrix(OTO)`} = I_2$
   
   
***   


However, the above items do not include **all** matrices for which the transpose commutes under multiplication.    

By the **Spectral Theorem**, the full set of matrices for which the transpose commutes are called **Normal** Matrices.  These are discussed in the text book in sections NM and OD (pages 574-580.)

By **Theorem OD (Orthonormal Diagonalization)** on page 575, we have:   

Suppose that $A$ is a square matrix. Then there is a **unitary** matrix $U$ and a **diagonal** matrix $D$, *with diagonal entries equal to the eigenvalues of* $A$, such that $U^TAU = D$ if and only if $A$ is a **normal** matrix.   

(Note: for matrices with **all real** entries, **unitary** is synonymous with **orthonormal** , and above I have used $U^T$ to indicate **transpose**, which for real matrices is the same as the **adjoint.**  The distinction only comes into play in the case of matrices with **complex** entries.)

### NORMAL example (non-symmetric, etc.):

```{r normal-matrices,eval=T}

Z = c(
  1,1,0,
  0,1,1,
  1,0,1)
Z = matrix(Z,3,3,T)
Z
# Compute transpose
ZT=t(Z)
ZT
# Compute ZZT
ZZT = Z %*% ZT
ZZT
# Compute ZTZ
ZTZ = ZT %*% Z
ZTZ
# Are they equal?
ZZT==ZTZ

```

For example, consider a **normal** matrix which is neither symmetric nor orthogonal:    


$$ Z = `r writeMatrix(Z)` ; Z^{T} = `r writeMatrix(ZT)`$$
$$ ZZ^{T} = `r writeMatrix(Z)` `r writeMatrix(ZT)` = `r writeMatrix(ZZT)`$$
$$ Z^{T}Z = `r writeMatrix(ZT)` `r writeMatrix(Z)` = `r writeMatrix(ZTZ)`$$


***

# Part 2 - Matrix Decomposition - LU

#### Matrix factorization is a very important problem. 
There are supercomputers built just to do matrix factorizations. 
Every second you are on an airplane, matrices are being factorized. 
Radars that track flights use a technique called Kalman filtering. 
At the heart of Kalman Filtering is a Matrix Factorization operation. 
Kalman Filters are solving linear systems of equations when they track your flight using radars. 

## Write an R function to factorize a square matrix A into LU or LDU, whichever you prefer. 

Please submit your response in an R Markdown document using our class naming convention, 
e.g. LFulton_Assignment2_PS2.Rmd
You don’t have to worry about permuting rows of A and you can assume that A is less than 5x5, if you need to hard-code any variables in your code. 
If you doing the entire assignment in R, then please submit only one markdown document for both the problems.    

### MYLU function:
```{r MYLU}
library(pracma)                       # to get functions like eye(), zeros(), ones()
debug=FALSE

MYLU = function (A) {

  n <- nrow(A);                       # how many rows (and, columns) are in the (square) matrix?
  L <- eye(n)                         # start L from the identity matrix
  U <- eye(n)                         # start U from the identity matrix
  
  for (k in 1:n) {
    if(debug) print(paste0("for k: ",k))
    U[k,k] <- A[k,k];
    for (i in (k+1):n) {
      if (i>n) break
      if(debug) print(paste0("(k,i): (", k,",",i,")"))
      L[i,k] <- (A[i,k])/(U[k,k]);
      U[k,i] <-  A[k,i]  
    }
    for (i in ((k+1):n)) {
      if (i>n) break
      if(debug) print(paste0("(k,i): (", k,",",i,")"))
      for (j in ((k+1):n)) {
        if (j>n) break
        if(debug) print(paste0("(k,i,j): (", k,",",i,",",j,")"))
        A[i,j] <- A[i,j] - L[i,k] %*% U[k,j];
      }
    }
  }
      
  myresult <- list(L,U)
  names(myresult) <- c("L","U")
  
  return(myresult)

} # end function MYLU
```


### Test a constructed 3x3 matrix

```{r test-MYLU1, eval=T}
# Create an upper triangular matrix
testU1 <- c(
  1,1,1,
  0,1,1,
  0,0,1)
testU1 <- matrix(testU1,3,3,T) 
testU1
det(testU1)
inv(testU1)

#Create a lower triangular matrix
testL1 <- c(
  1,0,0,
  1,1,0,
  1,1,1
  )
testL1 <- matrix(testL1,3,3,T)
testL1
det(testL1)
inv(testL1)

# Multiply together to get a matrix for testing
testLU1 <- testL1 %*% testU1
testLU1

```

Testing matrix ${`r writeMatrix(testLU1)`}$  :

#### Get the result using my LU algorithm

```{r test-MYLU1a, eval=T}
myresult1 <- MYLU(testLU1)
myresult1
```

My result is $$L = `r writeMatrix(myresult1$L)`  ; U = `r writeMatrix(myresult1$U)` ; LU = `r writeMatrix(myresult1$L %*% myresult1$U)` $$

```{r test-MYLU1b, eval=T}
# do my L and U match original L and U ?
myresult1$L == testL1
myresult1$U == testU1
```

#### Check using lu decomposition function from pracma
```{r test-MYLU1c}
library(pracma)
theirresult1 <- pracma::lu(testLU1)
theirresult1

myresult1
```

Pracma result is $$L = `r writeMatrix(theirresult1$L)`  ; U = `r writeMatrix(theirresult1$U)`; LU = `r writeMatrix(theirresult1$L %*% theirresult1$U)` $$

```{r test-MYLU1d, eval=T}


# Does my result match their result?
myresult1$L == theirresult1$L
myresult1$U == theirresult1$U
```


### Test a constructed 5x5 matrix


```{r test-MYLU2, eval=T}
# Create an upper triangular matrix
testU2 <- c(
  1,2,3,4,5,
  0,2,3,4,5,
  0,0,3,4,5,
  0,0,0,4,5,
  0,0,0,0,5
  )
testU2 <- matrix(testU2,5,5,T) 
testU2
det(testU2)
inv(testU2)

#Create a lower triangular matrix
testL2 <- c(
  1,0,0,0,0,
  1,1,0,0,0,
  1,1,1,0,0,
  1,1,1,1,0,
  1,1,1,1,1
  )
testL2 <- matrix(testL2,5,5,T)
testL2
det(testL2)
inv(testL2)

# Multiply together to get a matrix for testing
testLU2 <- testL2 %*% testU2
testLU2

```

Testing matrix ${`r writeMatrix(testLU2)`}$  :

#### Get the result using my LU algorithm

```{r test-MYLU2a, eval=T}
# Get the result using my LU algorithm
myresult2 <- MYLU(testLU2)
myresult2
```

My result is $$L = `r writeMatrix(myresult2$L)`  ; U = `r writeMatrix(myresult2$U)`; LU = `r writeMatrix(myresult2$L %*% myresult2$U)` $$

```{r test-MYLU2b, eval=T}

# do my L and U match original L and U ?
myresult2$L == testL2
myresult2$U == testU2


# Check using lu decomposition function from pracma
library(pracma)
theirresult2 <- pracma::lu(testLU2)
theirresult2

```

Pracma result is $$L = `r writeMatrix(theirresult2$L)`  ; U = `r writeMatrix(theirresult2$U)`; LU = `r writeMatrix(theirresult2$L %*% theirresult2$U)` $$

```{r test-MYLU2d, eval=T}


# Does my result match their result?
myresult2$L == theirresult2$L
myresult2$U == theirresult2$U
```





### Test a different 5x5 matrix


```{r test-MYLU3, eval=T}
# Create an upper triangular matrix
testU3 <- c(
  9,3,2,4,1,
  0,6,9,-3,5,
  0,0,4,2,-1,
  0,0,0,7,5,
  0,0,0,0,-12
  )
testU3 <- matrix(testU3,5,5,T) 
testU3
det(testU3)
inv(testU3)

#Create a lower triangular matrix
testL3 <- c(
  1,0,0,0,0,
  -1,1,0,0,0,
  99,5,1,0,0,
  101,22,-66,1,0,
  77,-55,33,-22,1
  )
testL3 <- matrix(testL3,5,5,T)
testL3
det(testL3)
inv(testL3)

# Multiply together to get a matrix for testing
testLU3 <- testL3 %*% testU3
testLU3

```

Testing matrix ${`r writeMatrix(testLU3)`}$  :

#### Get the result using my LU algorithm

```{r test-MYLU3a, eval=T}
# Get the result using my LU algorithm
myresult3 <- MYLU(testLU3)
myresult3
```

My result is $$L = `r writeMatrix(myresult3$L)`  ; U = `r writeMatrix(myresult3$U)` ; LU = `r writeMatrix(myresult3$L %*% myresult3$U)` $$

```{r test-MYLU3b, eval=T}


# do my L and U match original L and U ?
myresult3$L == testL3
myresult3$U == testU3


# Check using lu decomposition function from pracma
library(pracma)
theirresult3 <- pracma::lu(testLU3)
theirresult3

```

Pracma result is $$L = `r writeMatrix(theirresult3$L)`  ; U = `r writeMatrix(theirresult3$U)` ; LU = `r writeMatrix(theirresult3$L %*% theirresult3$U)` $$

```{r test-MYLU3d, eval=T}

# Does my result match their result?
myresult3$L == theirresult3$L
myresult3$U == theirresult3$U
```


#### In each of the examples tested, MYLU gives the same results as the LU function from the pracma package.
#### Such results successfully decompose the tested matrix back into the L and U constituents from which it was constructed.
