---
title: "605-HW03-Eigenvectors"
author: "Michael Y."
date: "September 15, 2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    code_folding: show
    toc: yes
    toc_float: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
classoption: portrait
editor_options:
  chunk_output_type: inline
---
<style>
  .main-container {
    max-width: 1200px !important;
  }
</style>---

#### Setup
```{r setup, eval=T, echo=F}
knitr::opts_chunk$set(echo = TRUE)
directory = "C:/Users/Michael/Dropbox/priv/CUNY/MSDS/201909-Fall/DATA605_Larry/20190915_Week03/"
knitr::opts_knit$set(root.dir = directory)

### Make the output wide enough
options(scipen = 999, digits=6, width=120)

### Load some libraries
library(tidyr)
library(dplyr)
library(kableExtra)
library(pracma)

```

``` {r writemtx, eval=T, echo=F}
#### Function to write Matrix, courtesy of Vinayak Patel :
writeMatrix <- function(x) {  
  begin <- "\\begin{bmatrix}"  
  end   <- "\\end{bmatrix}"  
  X     <-    apply(x, 1, function(x) {
      paste(
          paste(x, collapse = "&"),
          "\\\\"
      )
    }
  )  
  paste(c(begin, X, end), 
        collapse = "")
  }
```

\newpage
## Part 1 - Matrix Rank

### (1) What is the rank of the matrix A?
```{r rankA, eval=T, echo=F}
A = c(
 1, 2,  3,  4,
-1, 0,  1,  3, 
 0, 1, -2,  1,
 5, 4, -2, -3
)
A=matrix(A,nrow=4,byrow=T)
A
```
$$ A = `r writeMatrix(A)` $$
```{r is-A-full-rank, eval=T}
#B has a non-zero determinant, so it is non-singular, invertible, and full-rank:
det(A)

# All 4 rows of the Row-Reduced Echelon Form are non-zero:
rref(A)

# The calculated matrix rank:
library(matrixcalc)
matrix.rank(A)
```

Because ${A}$ is a square matrix with non-zero determinant, it is full rank, i.e., **rank = 4** .   
This is confirmed as the Row-Reduced Echelon Form contains 4 non-zero rows.  

### (2) Given an mxn matrix where m > n, what can be the maximum rank? 

The maximum rank is ${maxrank = min(m,n)}$ , which in this case is the number of columns, ${n}$, as there are more rows than columns.

#### The minimum rank, assuming that the matrix is non-zero?

The minimum rank is 1.


### (3) What is the rank of matrix ${B}$ ?  

``` {r part3, eval-T, echo=F}
B =
c(
1, 2, 1,
3, 6, 3,
2, 4, 2
)
B = matrix(B, nrow=3,byrow=T)
B
```


$$ B = `r writeMatrix(B)` $$

```{r checkB}
#B has a zero determinant, so it is singular, non-invertible, and less than full-rank.
det(B)

# Row-Reduced echolon form shows a single non-zero row:
rref(B)

# calculated rank is 1:
matrix.rank(B)

Bcol1 = as.matrix(B[,1])
```

B has only a single independent column:    ${ `r writeMatrix(Bcol1)`}$  .    
The third column is the same as the first, and the middle column is double the first.    
Also, the number of non-zero rows in the row-echelon form is 1.  
**Therefore, the matrix ${B}$ has rank 1.**   


***
\newpage
## Part 2 - Eigenvalues and Eigenvectors

### Compute the eigenvalues and eigenvectors of the matrix ${A}$ .     

You'll need to show your work.   
You'll need to write out the characteristic polynomial and show your solution.   

```{r 2matrixA, eval=T}
A =
c(
1, 2, 3,
0, 4, 5,
0, 0, 6
)
A = matrix(A,nrow=3,byrow=T)
A
```

$$ A = `r writeMatrix(A)` $$ 
$$ Ax = \lambda x $$
$$ Ax - \lambda x = 0 $$
$$ \left(A-\lambda I \right)x = 0 $$

$$ `r writeMatrix(A)` - \begin{bmatrix} \lambda  & 0 & 0 \\ 0 & \lambda  & 0 \\ 0 & 0 & \lambda  \end{bmatrix}  = 0 $$
$$ \begin{bmatrix} 1-\lambda  & 2 & 3 \\ 0 & 4-\lambda  & 5 \\ 0 & 0 & 6-\lambda  \end{bmatrix} x = 0 $$

$$ det \begin{bmatrix} 1-\lambda  & 2 & 3 \\ 0 & 4-\lambda  & 5 \\ 0 & 0 & 6-\lambda  \end{bmatrix} = 0 $$
Because the matrix is upper triangular, the determinant equals the product of the diagonal entries:

$$ (1-\lambda) (4-\lambda)(6-\lambda) = 0  $$
The characteristic polynomial is ${24-34x+11x^2-x^3}=0$, but we can easily see from the factored form above that the eigenvalues are ${\lambda \in \{6,4,1\}}$ .

\newpage

### Eigenvector for $\lambda=6$ :

$$ \begin{bmatrix} 
1-\lambda  & 2 & 3 \\ 
0 & 4-\lambda  & 5 \\ 
0 & 0 & 6-\lambda  
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\lambda \begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}\\ $$

$$ \begin{bmatrix} 
1-6 & 2 & 3 \\ 
0 & 4-6 & 5 \\ 
0 & 0 & 6-6 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\begin{bmatrix} 
-5 & 2 & 3 \\ 
0 & -2 & 5 \\ 
0 & 0 & 0 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}=0 $$


$$ \begin{cases} -5x_{ 1 }+2x_{ 2 }+3x_{ 3 }=0 \\ 0x_{ 1 }-2x_{ 2 }+5x_{ 3 }=0 \\ 0x_{ 1 }+0x_{ 2 }+0x_{ 3 }=0 \end{cases}\\ $$

$$ \Rightarrow \begin{cases} 
-5x_{ 1 }+5x_{ 3 }+3x_{ 3 }=0 \\ 
5x_{ 3 }=2x_{ 2 } 
\end{cases}\\
\Rightarrow 
\begin{cases} 
x_{ 2 }=\frac { 5x_{ 3 } }{ 2 }  \\ 
5x_{ 1 }=8x_{ 3 } 
\end{cases}\\
\Rightarrow 
\begin{cases} 
x_{ 2 }=\frac { 5x_{ 3 } }{ 2 } =2.5(x_{ 3 }) \\ 
x_{ 1 }=\frac { 8 }{ 5 } 
x_{ 3 }=1.6(x_{ 3 }) 
\end{cases}\\  
\Rightarrow 
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}=\begin{bmatrix} 1.6 \\2.5 \\ 1.0 \end{bmatrix}\\\\ $$

Check:
```{r check-eig6,eval=T}
# Eigenvector for lambda = 6
x6 = c(1.6,2.5,1)
x6

# Ax / 6
result6 = (A %*% x6) / 6
result6

# Ax - 6x
result6 - x6

# Is this result zero?
result6 - x6 < 1e-10

# Normalize this result
sumsq_x6 = sum(x6^2)
sumsq_x6
x6_norm = x6 / sqrt(sumsq_x6)
x6_norm
```

#### Normalized Eigenvector for lambda = 6:

$$ \begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\begin{bmatrix} 1.6 \\2.5 \\ 1.0 \end{bmatrix} \cdot \frac{1}{\sqrt{9.81}}
=\begin{bmatrix} .510841 \\ .798189 \\ .319275 \end{bmatrix} $$



\newpage

### Eigenvector for $\lambda=4$ :

$$ \begin{bmatrix} 
1-4 & 2 & 3 \\ 
0 & 4-4 & 5 \\ 
0 & 0 & 6-4 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\begin{bmatrix} 
-3 & 2 & 3 \\ 
0 & 0 & 5 \\ 
0 & 0 & 2 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}=0 $$

$$ \begin{cases} 
-3x_{ 1 }+2x_{ 2 }+3x_{ 3 }=0 \\ 
0x_{ 1 }+0x_{ 2 }+5x_{ 3 }=0 \\ 
0x_{ 1 }+0x_{ 2 }+2x_{ 3 }=0 
\end{cases} 
\Rightarrow 
\begin{cases} 
x_{ 3 }=0 \\
-3x_{ 1 }+2x_{ 2 }=0 
\end{cases} 
\Rightarrow 
\begin{cases} 
x_{ 3 }=0 \\
3x_{ 1 }=2x_{ 2 } 
\end{cases}
\Rightarrow 
\begin{cases} 
x_{ 3 }=0 \\
x_{ 1 }=\frac{2x_{ 2 }}{3} 
\end{cases}\\
\Rightarrow
\begin{bmatrix} 
x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } 
\end{bmatrix}
=\begin{bmatrix} 2 \\3 \\ 0 
\end{bmatrix}\\\\ $$

```{r check-eig4,eval=T}
# Eigenvector for lambda = 4
x4 = c(2,3,0)
x4

# Ax / 4
result4 = (A %*% x4) / 4
result4

# Ax - 4x
result4 - x4

# Is this result zero?
result4 - x4 < 1e-10

# Normalize this result
sumsq_x4 = sum(x4^2)
sumsq_x4
x4_norm = x4 / sqrt(sumsq_x4)
x4_norm
```


#### Normalized Eigenvector for lambda = 4:

$$ \begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\begin{bmatrix} 2 \\3 \\ 0 \end{bmatrix} \cdot \frac{1}{\sqrt{13}}
=\begin{bmatrix} .55470 \\ .83205 \\ 0 \end{bmatrix} $$




\newpage

### Eigenvector for $\lambda=1$ :

$$ \begin{bmatrix} 
1-1 & 2 & 3 \\ 
0 & 4-1 & 5 \\ 
0 & 0 & 6-1 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}
=\begin{bmatrix} 
0 & 2 & 3 \\ 
0 & 3 & 5 \\ 
0 & 0 & 5 
\end{bmatrix}
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}=0 $$

$$ \begin{cases} 
0x_{ 1 }+2x_{ 2 }+3x_{ 3 }=0 \\ 
0x_{ 1 }+3x_{ 2 }+5x_{ 3 }=0 \\ 
0x_{ 1 }+0x_{ 2 }+5x_{ 3 }=0 
\end{cases} 
\Rightarrow 
\begin{cases} 
x_{ 3 }=0 \\
x_{ 2 }=0 \\
x_{ 1 }=1 
\end{cases}\\ 
\Rightarrow
\begin{bmatrix} x_{ 1 } \\ x_{ 2 } \\ x_{ 3 } \end{bmatrix}=\begin{bmatrix} 1 \\0 \\ 0 \end{bmatrix}\\\\ $$

Check:

```{r check-eig1,eval=T}
# Eigenvector for lambda = 1
x1 = c(1,0,0)
x1

# Ax / 1
result1 = (A %*% x1) / 1
result1

# Ax - 1x
result1 - x1

# Is this result zero?
result1 - x1 < 1e-10

# Normalize this result
sumsq_x1 = sum(x1^2)
sumsq_x1
x1_norm = x1 / sqrt(sumsq_x1)
x1_norm
```

#### My eigenvectors

```{r myeigvecs, eval=T}
myeigvecs = cbind(x6_norm, x4_norm, x1_norm)
myeigvecs
```

My normalized eigenvectors (each column):  ${`r writeMatrix(myeigvecs)`}$

\newpage

### Check results vs. eigen() function

```{r check-Eigen, eval=T}
result = eigen(A)
eigvals = result$values
eigvecs = result$vectors
```

$$ Eigenvalues:  \lambda = `r writeMatrix(as.matrix(eigvals))` $$
$$ Eigenvectors: x = `r writeMatrix(as.matrix(eigvecs))` $$

```{r check-Eigen2, eval=T}
# Check if Ax[i] = lambda[i] x[i]
lhs = rhs = difference = rep.int(list(NULL),3)
lhstext = rhstext = difftext = rep.int(list(NULL),3)

#whicheig = 1
for (whicheig in 1:3) {

lhstext[[whicheig]] = paste0("A x_{",whicheig,"}")  
lhs[[whicheig]] = A %*% eigvecs[,whicheig]

rhstext[[whicheig]] = paste0(eigvals[whicheig]," x_{",whicheig,"}")  
rhs[[whicheig]] = as.matrix(eigvals[whicheig] * eigvecs[,whicheig])

difference[[whicheig]] = lhs[[whicheig]] - rhs[[whicheig]]

}
```
$$ Checking \quad A  x_{1} = \lambda_{1}  x_{1}  \quad : \quad  \lambda_{1} = `r eigvals[1]` \quad ; \quad x_{1} = `r writeMatrix(as.matrix(eigvecs[,1]))` $$
$$ `r lhstext[[1]]`  =  `r writeMatrix(lhs[[1]])` \quad ; \quad `r rhstext[[1]]`  =  `r writeMatrix(rhs[[1]])`  \quad ; \quad diff = `r writeMatrix(difference[[1]])` $$

$$ Checking \quad A x_{2} = \lambda_{2} x_{2}   \quad : \quad  \lambda_{2} = `r eigvals[2]` \quad ; \quad x_{2} = `r writeMatrix(as.matrix(eigvecs[,2]))` $$
$$ `r lhstext[[2]]`  =  `r writeMatrix(lhs[[2]])` \quad ; \quad `r rhstext[[2]]`  =  `r writeMatrix(rhs[[2]])`  \quad ; \quad diff = `r writeMatrix(difference[[2]])` $$

$$ Checking \quad A x_{3} = \lambda_{3} x_{3}  \quad : \quad  \lambda_{3} = `r eigvals[3]` \quad ; \quad x_{3} = `r writeMatrix(as.matrix(eigvecs[,3]))` $$
$$ `r lhstext[[3]]`  =  `r writeMatrix(lhs[[3]])` \quad ; \quad `r rhstext[[3]]`  =  `r writeMatrix(rhs[[3]])`  \quad ; \quad diff = `r writeMatrix(difference[[3]])` $$

#### Comparing my eigenvectors with those from eigen() :
```{r finalcompare, eval=T}
### my eigenvectors
myeigvecs

### eigvecs from eigen()
eigvecs

### difference
myeigvecs - eigvecs

### Equal to zero?
myeigvecs - eigvecs < 1e-10
```


Please show your work using an R-markdown document. 
Please name your assignment submission with your first initial and last name.

